Higher-Order unification for free!

### 2024-04-17 paper/main.tex
update progress eta

\newcommand{\progressetaleft}{\emph{$\eta$-progress-\lhs}\xspace}
\begin{definition}[\progressetaleft]\label{def:progressetaleft}
A link \linketaM{\Gamma}{X}{T} is removed from \linkStore when
$X$ becomes rigid. Let $y\in\Gamma$, there are two cases:
\begin{enumerate}
  \item if $X = a$ or $X = y$ or $X = f \appsep a_1\ldots a_n$
    we unify the $\eta$-expansion of $X$ with $T$, that is we run
    $\lambda x.X \appsep x\Ue{} T$
    \item if $X = \lambda x.t$ we run $X \Ue{} T$.
\end{enumerate}
\end{definition}

\newcommand{\progressetaright}{\emph{$\eta$-progress-\rhs}\xspace}
\begin{definition}[\progressetaright]\label{def:progressetaright} A link
\linketaM{\Gamma}{X}{T} is removed from \linkStore when either 1)
$\testmaybeeta~T$ does not hold (anymore) or 2) by $\eta$-contracting $T$ to
$T'$ where $T'$ does not start with the \elpiIn{lam} constructor. In the
first case, $X$ is unified with $T$, and in the second one, $X$ is unified with
$T'$ (under the context $\Gamma$).
\end{definition}

There is a third case in which a link is removed from \linkStore, namely
when the \lhs is assigned to a variable that is the \lhs of another
\linketa.

@@

\begin{lemma}
  Given a \linketa \linketaM{\Gamma}{X}{\lambda x.t}, the unification done by \progressetaright is between
  terms in \wellb.
  \label{lemma:unif-eta-2}
\end{lemma}

\begin{proof}[Proof sketch]
  $X$ is variable, and, by \cref{def:progressetaright}, $\lambda x.t$ is either no more
  a \maybeeta, i.e. it is not a $\eta$-expansion and, so, $\wellb(\lambda x.t)$, 
  the unification is performed between $X$ and $\lambda x.t$.
  Otherwise, $\lambda x.t$ has the shape $\lambda x.t' \appsep x$ with $x$ not free in $t'$ and
   $\wellb(t')$. In this second case the unification is done between $X$ and $t'$.
  In both cases, unifications are between \wellb terms.
\end{proof}

@@

\begin{proof}[Proof sketch]
  By \cref{lemma:unif-eta-1,lemma:unif-eta-2,lemma:unif-eta-3}, every
  unification performed by the activation of a \linketa is done between
  terms in \wellb, therefore, the substitution remains \wellb.
\end{proof}

@@

\begin{proof}[Proof sketch]
  Rules \cref{def:progressetaleft,def:progressetaright} and
  \cref{def:progressetadedup} remove one link from \linkStore, hence they
  cannot be applied indefinitely.
  Moreover each rule only relies on terminating operations such as \Ue,
  $\eta$-contraction, $\eta$-expansion, relocation (a recursive copy of a
  finite term).
  % The addition of rules for \elpiIn{progres1} complicates the function
  % \elpiIn{progress}. We can note, however, that they do not prevent the
  % termination of \elpiIn{progress}. 1) If a link is activated it is removed from
  % \linkStore and the recursive call to \elpiIn{progress} will have a smaller
  % list of links to recurse on. Moreover, link activation only runs terminating
  % instructions (such as unification). 2) If a link is deduplicated, the
  % termination of \elpiIn{progress} is still guaranteed since again we reduce
  % \linkStore and the instructions run by link deduplications are all
  % terminating. 3) If a link is neither activated nor deduplicated, i.e. it
  % remains suspended, then \linkStore remains unchanged like the substitution;
  % therefore, \elpiIn{if (L = L1, S1 = S2)} succeeds and \elpiIn{progress}
  % terminates.
\end{proof}

@@

\begin{proof}[Proof sketch]\todo{to reformulate}
  % For any step $i$, we want to prove that $\fstep_i$ succeeds iff $\hstep_i$
  % succeeds. A $\hstep_i$ is made by a unification
  % $u = \sigma_{i-1}\hoUnifPb_{i_l} \Ue \sigma_{i-1}\hoUnifPb_{i_r} \mapsto \sigma_i'$ and a link
  % progression $p = \progress(\linkStore_{i-1},\sigma_i') \mapsto (\linkStore_i,\sigma_i)$.
  For any step $i$, we want to prove that $\fstep_i$ succeeds iff $\hstep_i$
  succeeds. A $\hstep_i$ is made by a unification $u$ and a link
  progression $p$.
  Without loss of generality,\todo{explain why} we consider a \Fo unification problem of the form $s_1 \Uo s_2$
  where $s_1 \in \maybeeta$, which is compiled into the \linketa $\linketaM{}{X}{\lambda x.t_1}$. 
  The unification $u$ always succeeds, since $X$ is a fresh variable, we 
  only have to treat the progression $p$.
  Two cases should be analyzed:
  \begin{itemize}
    \item $s_2 \notin \maybeeta$, in this case, the unification problem becomes
          $X \Ue t_2$, which assigns $X$ to $t_2$.
          After the unification, \progressetaleft triggers the \linketa,
          and $\lambda x.t_1$ is unified with $t_2$ (or its $\eta$-expansion when needed).
          This unification succeeds iff $s_1$ and $s_2$ unify. \progressetaleft
          aims essentially to transpose the two rules \ref{clause:eta1} and \ref{clause:eta2}
          of the object language in the meta one.
    \item $s_2 \in \maybeeta$, in this case, the unification problem becomes $X \Ue Y$ 
          with the extra link $\linketaM{}{Y}{\lambda x.t_2}$. After the unification of 
          $X$ and $Y$, \progressetadedup is fired and $t_1$ and $t_2$ are unified.
          This unification succeeds iff $s_1 \Uo s_2$ succeeds.
  \end{itemize}
  In this setting, \progressetaright does not play a critical role since \mapStore
  is a bijection; it's role is to perform variable refinement, i.e. it never fails.
  % wlog we assume s1 is in \maybeeta, s2 is not (if both are thanks to dedup we are back in this situation).
  % we assume s1 is compiled to X and $\linketaM{\Gamma}{X}{t1}$.
  % In all cases $X \Ue t2$ succeeds since X is a variable and
  % hence assigns t2 to X.
  % Rule \progressetaleft unify with \Ue X (hence t2) with t1, with the only
  % additional care of eta expanding t2 if needed (5.9.1).
% 
  % Two cases should be analyzed: 1) $\fstep_i$ succeeds to unify 
  % $s1$ with $s2$. The interesting case is when fostep succeds by
  % finding a $\rho$ such that $\rho s1 \Eo \rho s2$ requires
  % rules~\ref{clause:eta1} and \ref{clause:eta2} at the top level,
  % and that is in exactly the cases in which T (from page 3) is
  % not a lam (see 5.9.1) compensating for the missing rules in \Ue.
  % 
  % 2) If $\fstep_i$ fails to unify $s1 \Uo s2$
  % it means that, after applying rules~\ref{clause:eta1} and \ref{clause:eta2}
  % and the one flam-flam we find different terms.
  % (5.9.1) only adds lambdas that are forcibly consumed by
  % lam-lam rule.
\end{proof}

@@

Note \progressetaright is only needed in the next section where chains of links
are added. Hence we give an example for rule \progressetaright in the next section.
\todo{fix english}

@@

>>>>

\newcommand{\progressetaleft}{\emph{$\eta$-progress-\lhs}\xspace}
\begin{definition}[\progressetaleft]\label{def:progressetaleft}
A link \linketaM{\Gamma}{X}{t} is removed from \linkStore when
$X$ becomes rigid. Let $y\in\Gamma$, there are two cases:
\begin{enumerate}
  \item if $X = a$ or $X = y$ or $X = f \appsep a_1\ldots a_n$
    we unify the $\eta$-expansion of $X$ with $t$, that is we run
    $\lambda x.X \appsep x\Ue{} t$
    \item if $X = \lambda x.t$ we run $X \Ue{} t$.
\end{enumerate}
\end{definition}

\newcommand{\progressetashrink}{\emph{$\eta$-progress-shrink}\xspace}
\begin{definition}[\progressetashrink]\label{def:progressetashrink}
Given a link \linketaM{\Gamma}{X}{t}, and let $S$ the scope of $X$
we pose $T = S \cap \Gamma$. If $T$ is not empty,
then the original link is replaced by a new link
\linketaM{S}{X}{t'}, where $t'$ is a term where all the variables in $\Gamma / X$ 
have been pruned.
\end{definition}

% There is a third case in which a link is removed from \linkStore, namely
% when the \lhs is assigned to a variable that is the \lhs of another
% \linketa.

@@

% \begin{lemma}
%   Given a \linketa \linketaM{\Gamma}{X}{\lambda x.t}, the unification done by \progressetaright is between
%   terms in \wellb.
%   \label{lemma:unif-eta-2}
% \end{lemma}

% \begin{proof}[Proof sketch]
%   $X$ is variable, and, by \cref{def:progressetaright}, $\lambda x.t$ is either no more
%   a \maybeeta, i.e. it is not a $\eta$-expansion and, so, $\wellb(\lambda x.t)$, 
%   the unification is performed between $X$ and $\lambda x.t$.
%   Otherwise, $\lambda x.t$ has the shape $\lambda x.t' \appsep x$ with $x$ not free in $t'$ and
%    $\wellb(t')$. In this second case the unification is done between $X$ and $t'$.
%   In both cases, unifications are between \wellb terms.
% \end{proof}

@@

\begin{proof}[Proof sketch]
  By \cref{lemma:unif-eta-1,lemma:unif-eta-3}, every
  unification performed by the activation of a \linketa is done between
  terms in \wellb, therefore, the substitution remains \wellb.
\end{proof}

@@

\begin{proof}[Proof sketch]
  Rules \cref{def:progressetaleft} and
  \cref{def:progressetadedup} remove one link from \linkStore, hence they
  cannot be applied indefinitely.
  Moreover each rule only relies on terminating operations such as \Ue,
  $\eta$-contraction, $\eta$-expansion, relocation (a recursive copy of a
  finite term).
  % The addition of rules for \elpiIn{progres1} complicates the function
  % \elpiIn{progress}. We can note, however, that they do not prevent the
  % termination of \elpiIn{progress}. 1) If a link is activated it is removed from
  % \linkStore and the recursive call to \elpiIn{progress} will have a smaller
  % list of links to recurse on. Moreover, link activation only runs terminating
  % instructions (such as unification). 2) If a link is deduplicated, the
  % termination of \elpiIn{progress} is still guaranteed since again we reduce
  % \linkStore and the instructions run by link deduplications are all
  % terminating. 3) If a link is neither activated nor deduplicated, i.e. it
  % remains suspended, then \linkStore remains unchanged like the substitution;
  % therefore, \elpiIn{if (L = L1, S1 = S2)} succeeds and \elpiIn{progress}
  % terminates.
\end{proof}

@@

\begin{proof}[Proof sketch]\todo{to reformulate}
  % For any step $i$, we want to prove that $\fstep_i$ succeeds iff $\hstep_i$
  % succeeds. A $\hstep_i$ is made by a unification
  % $u = \sigma_{i-1}\hoUnifPb_{i_l} \Ue \sigma_{i-1}\hoUnifPb_{i_r} \mapsto \sigma_i'$ and a link
  % progression $p = \progress(\linkStore_{i-1},\sigma_i') \mapsto (\linkStore_i,\sigma_i)$.
  For any step $i$, we want to prove that $\fstep_i$ succeeds iff $\hstep_i$
  succeeds. A $\hstep_i$ is made by a unification $u$ and a link
  progression $p$.
  Without loss of generality,\todo{explain why} we consider a \Fo unification problem of the form $s_1 \Uo s_2$
  where $s_1 \in \maybeeta$, which is compiled into the \linketa $\linketaM{}{X}{\lambda x.t_1}$. 
  The unification $u$ always succeeds, since $X$ is a fresh variable, we 
  only have to treat the progression $p$.
  Two cases should be analyzed:
  \begin{itemize}
    \item $s_2 \notin \maybeeta$, in this case, the unification problem becomes
          $X \Ue t_2$, which assigns $X$ to $t_2$.
          After the unification, \progressetaleft triggers the \linketa,
          and $\lambda x.t_1$ is unified with $t_2$ (or its $\eta$-expansion when needed).
          This unification succeeds iff $s_1$ and $s_2$ unify. \progressetaleft
          aims essentially to transpose the two rules \ref{clause:eta1} and \ref{clause:eta2}
          of the object language in the meta one.
    \item $s_2 \in \maybeeta$, in this case, the unification problem becomes $X \Ue Y$ 
          with the extra link $\linketaM{}{Y}{\lambda x.t_2}$. After the unification of 
          $X$ and $Y$, \progressetadedup is fired and $t_1$ and $t_2$ are unified.
          This unification succeeds iff $s_1 \Uo s_2$ succeeds.
  \end{itemize}
  % wlog we assume s1 is in \maybeeta, s2 is not (if both are thanks to dedup we are back in this situation).
  % we assume s1 is compiled to X and $\linketaM{\Gamma}{X}{t1}$.
  % In all cases $X \Ue t2$ succeeds since X is a variable and
  % hence assigns t2 to X.
  % Rule \progressetaleft unify with \Ue X (hence t2) with t1, with the only
  % additional care of eta expanding t2 if needed (5.9.1).
% 
  % Two cases should be analyzed: 1) $\fstep_i$ succeeds to unify 
  % $s1$ with $s2$. The interesting case is when fostep succeds by
  % finding a $\rho$ such that $\rho s1 \Eo \rho s2$ requires
  % rules~\ref{clause:eta1} and \ref{clause:eta2} at the top level,
  % and that is in exactly the cases in which T (from page 3) is
  % not a lam (see 5.9.1) compensating for the missing rules in \Ue.
  % 
  % 2) If $\fstep_i$ fails to unify $s1 \Uo s2$
  % it means that, after applying rules~\ref{clause:eta1} and \ref{clause:eta2}
  % and the one flam-flam we find different terms.
  % (5.9.1) only adds lambdas that are forcibly consumed by
  % lam-lam rule.
\end{proof}

@@

@@

In order to grant fidelity, we add a new link progression for \linketa{}s.

\newcommand{\progressetaright}{\emph{$\eta$-progress-\rhs}\xspace}
\begin{definition}[\progressetaright]\label{def:progressetaright} A link
\linketaM{\Gamma}{X}{t} is removed from \linkStore when either 1)
$\testmaybeeta~t$ does not hold (anymore) or 2) by $\eta$-contracting $t$ to
$t'$ where $t'$ does not start with the \elpiIn{lam} constructor. In the
first case, $X$ is unified with $t$, and in the second one, $X$ is unified with
$t'$ (under the context $\Gamma$).
\end{definition}
